---
title: "Tutorial"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = FALSE)
```


## Annotate genes from Biomart

```{r}
library(biomaRt)
# List marts 
marts<- useMart("plants_mart", host="plants.ensembl.org")

head(listDatasets(marts))
listMarts(host="plants.ensembl.org")


ensembl <- useDataset("stuberosum_eg_gene", marts)
ensembl

Available_attributes<-listAttributes(ensembl, page = "feature_page")$description
head(keys(ensembl, keytype = "ensembl_gene_id"))
Available_attributes
#write.csv(Available_attributes, "~/Documents/Projects/PotatoNGS/reults/Avaialable_annotations.csv",   row.names = FALSE)


values <- rownames(counts)
##attributes are whatever you want tor retrieve from database

Row_Anno<-getBM(attributes = c("ensembl_transcript_id","ensembl_gene_id","description", "chromosome_name","start_position","end_position"), filters = "ensembl_transcript_id", values = values, mart = ensembl, uniqueRows = FALSE)

