---
title: "Tutorial"
author: Ashfaq Ali
email: ashfaq.ali@nbis.se
output: learnr::tutorial
runtime: shiny_prerendered
date: "`r format(Sys.time(), '%d %B, %Y')`"
---

```{r setup, include=FALSE}
library(learnr, lib.loc =  "/Users/cob-aaf/opt/anaconda3/envs/omicsMeta/lib/R/library")
knitr::opts_chunk$set(echo = FALSE)
usethis::
```


```{r}

library(devtools, "/Users/cob-aaf/opt/anaconda3/envs/omicsMeta/lib/R/library")
#devtools::install_github("metaOmic/preproc")
devtools::install_github("metaOmics/MetaQC")
devtools::install_github("metaOmics/MetaDE")
devtools::install_github("metaOmics/MetaPath")
install_github("Caleb-Huo/MetaSparseKmeans", build_vignettes=TRUE)
install_github("Caleb-Huo/IS-Kmeans")
install_github("Caleb-Huo/AWFisher")
devtools::install_github("metaOmics/metaPCA")
devtools::install_github("metaOmics/MetaDCN")
```

```{r}
install.packages('abind')
install.packages('zoo')
install.packages('xts')
install.packages('quantmod')
install.packages('ROCR')
install.packages("DMwR")
```


```{r}
l <-c("MetaQC", "MetaDE", "MetaPath", "metaPCA", "MetaDCN", "MetaSparseKmeans", "metafor", "GeneMeta")
invisible( lapply(l, library, quietly = TRUE, character.only=TRUE) )

```


## QC of micro array data

```{r}
# Let us load some example microarray data

load("~/Documents/GitHub/meta-analyses/metaOmics/data/prostate8.rda")
load("~/Documents/GitHub/meta-analyses/metaOmics/data/pathwayQC.rda")

```


```{r}
data(prostate8)
data(pathway)
DList=prostate8$data
colLabel=prostate8$dataLabel
GList=pathway[[1]]
filterGenes=TRUE
cutRatioByMean=0.3
cutRatioByVar=0.3

QCresult=MetaQC(DList, colLabel, GList,filterGenes,cutRatioByMean,cutRatioByVar)
plotMetaQC(QCresult$scoreTable)```
```


```{r}
data('Leukemia')
data('LeukemiaLabel')
data <- Leukemia
K <- length(data)

clin.data <- lapply(label, function(x) {data.frame(x)} )
for (k in 1:length(clin.data)){
 colnames(clin.data[[k]]) <- "label"
}
select.group <- c('inv(16)','t(15;17)')
ref.level <- "inv(16)"
data.type <- "continuous"
ind.method <- c('limma','limma','sam')
resp.type <- "twoclass"
paired <- rep(FALSE,length(data))
meta.method <- "Fisher"
meta.res <- MetaDE(data=data,clin.data = clin.data,
                    data.type=data.type,resp.type = resp.type,
                    response='label',
                    ind.method=ind.method, meta.method=meta.method,
                    select.group = select.group, ref.level=ref.level,
                    paired=paired,tail='abs',parametric=TRUE)
meta.method <- "Fisher.OC"
meta.res <- MetaDE(data=data,clin.data = clin.data,
                    data.type=data.type,resp.type = resp.type,
                    response='label',
                    ind.method=ind.method, meta.method=meta.method,
                    select.group = select.group, ref.level=ref.level,
                    paired=paired,tail='high',parametric=FALSE,nperm=100)
meta.method <- "FEM"
meta.res <- MetaDE(data=data,clin.data = clin.data,
                    data.type=data.type,resp.type = resp.type,
                    response='label',
                    ind.method=ind.method, meta.method=meta.method,
                    select.group = select.group, ref.level=ref.level,
                    paired=paired, tail='abs')
meta.method <- "REM"
REM.type <- "HO"
meta.res <- MetaDE(data=data,clin.data = clin.data,
                    data.type=data.type,resp.type = resp.type,
                    response='label',
                    ind.method=ind.method, meta.method=meta.method,
                    select.group = select.group, ref.level=ref.level,
                    paired=paired, 
                    REM.type=REM.type,tail='abs')
meta.method <- "SR"
meta.res <- MetaDE(data=data,clin.data = clin.data,
                    data.type=data.type,resp.type = resp.type,
                    response='label',
                    ind.method=ind.method, meta.method=meta.method,
                    select.group = select.group, ref.level=ref.level,
                    paired=paired,tail='abs',parametric=FALSE,nperm=100)
meta.method <- 'minMCC'
meta.res <- MetaDE(data=data,clin.data = clin.data,
                    data.type=data.type,resp.type = resp.type,
                    response='label',
                    ind.method=ind.method, meta.method=meta.method,
                    select.group = select.group, ref.level=ref.level,
                    paired=paired,tail='abs',parametric=FALSE,nperm=100)
meta.method <- "AW"
meta.res <- MetaDE(data=data,clin.data = clin.data,
                    data.type=data.type,resp.type = resp.type,
                    response='label',covariate = NULL,
                    ind.method=ind.method, meta.method=meta.method,
                    select.group = select.group, ref.level=ref.level,
                    paired=paired, rth=NULL,
                    REM.type=NULL,tail='abs',parametric=TRUE)
```

If you get a Lang error, try following

```{r}
https://stackoverflow.com/questions/9689104/installing-r-on-mac-warning-messages-setting-lc-ctype-failed-using-c
```

```{r}

```

```{r}

```


```{r}
browseVignettes("MetaSparseKmeans")
```




```{r}
# install.packages(c('GSA', 'combinat', 'samr', 'survival', 'cluster', 'gplots', 'ggplot2', 'irr', 'shape', 'snow', 'snowfall', 'igraph', 'doMC', 'PMA'))
# 
# source('https://bioconductor.org/biocLite.R')
# BiocManager::install(c('multtest', 'Biobase', 'edgeR', 'DESeq2', 'impute', 'limma', 'AnnotationDbi', 'Consensus-ClusterPlus', 'genefilter', 'GSEABase', 'Rgraphviz'))
```



in terminal use following to clone the repository

```{bash}
git clone "https://github.com/metaOmic/metaOmics"
```

```{r}
install.packages("shiny")
install.packages("DMwR")
shiny::runApp('metaOmics', port=9987)

## Data
Import data from the 


### Exercise 

Here's a simple exercise with an empty code chunk provided for entering the answer.*

Write the R code required to add two plus two:
````


```{r two-plus-two, exercise=TRUE}
source("./metaOmics/global/constants.R")
source("./metaOmics/global/messages.R")
source("./metaOmics/global/help.R")
#source("./metaOmics/global/database.R")
source("./metaOmics/global/helpers.R")
source("./metaOmics/global/directoryInput.R")
```

```{r}
library(GeneMeta)
library(RColorBrewer)
> #load("~/Bioconductor/Projects/GraphCombine/MetaBreast/data/Nevins.RData") > data(Nevins)
  
```


### Exercise with Code

*Here's an exercise with some prepopulated code as well as `exercise.lines = 5` to provide a bit more initial room to work.*

Now write a function that adds any two numbers and then call it:

```{r add-function, exercise=TRUE, exercise.lines = 5}
add <- function() {
  
}
```

## Random effect models

### Exercise with Hint

*Here's an exercise where the chunk is pre-evaulated via the `exercise.eval` option (so the user can see the default output we'd like them to customize). We also add a "hint" to the correct solution via the chunk immediate below labeled `print-limit-hint`.*

Modify the following code to limit the number of rows printed to 5:

```{r print-limit, exercise=TRUE, exercise.eval=TRUE}
rmarkdown::metadata$title
```

```{r print-limit-hint}
head(mtcars)
```

### Quiz

*You can include any number of single or multiple choice questions as a quiz. Use the `question` function to define a question and the `quiz` function for grouping multiple questions together.*

Some questions to verify that you understand the purposes of various base and recommended R packages:

```{r quiz}
quiz(
  question("Which package contains functions for installing other R packages?",
    answer("base"),
    answer("tools"),
    answer("utils", correct = TRUE),
    answer("codetools")
  ),
  question("Which of the R packages listed below are used to create plots?",
    answer("lattice", correct = TRUE),
    answer("tools"),
    answer("stats"),
    answer("grid", correct = TRUE)
  )
)
```

## Fixed effect models

```{r}

```

### Exercise


## Meta analyses of gene expression data

### Gene meta package in R

## Network meta analyses